--- 
title: "Time series of ion concentrations in Puerto Rico streams before and after Hurricane Hugo (1989)"
format: html
---

![*Trajectory of Hurricane Hugo.* [Source](https://www.wbtw.com/news/grand-strand/remembering-hurricane-hugo-35-years-later/)](https://www.wbtw.com/wp-content/uploads/sites/22/2024/09/Screenshot-2024-09-21-080156.png?strip=1)



### Background 

Ion concentrations of potassium (K), nitrate (NO3-), magnesium (Mg), calcium (Ca), and ammonium (NH4+) in Puerto Rico Streams were collected from 1988-1984. After Hurricane Hugo hit Puerto Rico in Septermber of 1989, Shaefer et al. (2000) were interested in the hurricane's long-term impact on ion concentrations among four Puerto Rico streams in the Luquillo mountain region.  



### Data
Data is downloaded from [McDowell and International Institute Of Tropical Forestry (IITF) 2024](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064). 

### Methods

To replicate the figure, cleaning and wrangling were performed. Data were individually filtered for relvant dates and ion columns. Upon completion, the the 9-week moving average for each site were calculated and combined into one data frame. The ion columns were then put into into one column to facillate plotting. 


```{r}
#| output: false
# load package
library(tidyverse)
```

```{r}
#| output: false
# load cleaned data 
streams_pivot <- read_csv("../output/output.csv")
```




```{r}
# reorder the ions to match figure 3
streams_pivot$ion <- factor(
  streams_pivot$ion,
  levels = c("K", "NO3", "Mg", "Ca", "NH4")
)
```


Plot time on the x-axis, and mean ion concentration on the y-axis. The vertical line on September 17, 1989 to represent date of Hurricane Hugo event. 


```{r}
# plot
fig3 <- ggplot(
  streams_pivot,
  mapping = aes(
    x = window_start,
    y = mean_concentration,
    linetype = stream
  )
  ) +
  geom_line() +
  facet_wrap(~ion, scales = "free", ncol = 1, strip.position = "left" ) +
  labs(
    x = "Date",
    y = "Ion concentration",
    caption = "Concentrations of ions before and after Hurricane Hugo in four Puerto Rico streams. "
    ) +
  geom_vline(xintercept = ymd("1989-09-17"), color = "darkgrey") +
  theme_bw()
```


# Results

Print the result.

The graph shows the ion concentration in four Puerto Rico streams: Rio Mameyes Puente Roto (PRM), Bisley Quebradas 1, Bisley Quebradas 2, Bisley Quebradas 3. Specifically, the figure plots the concentrations of potassium (K mg l-1), nitrate (NO3 µg l-1), magnesium (mg l-1), calcium (mg l-1), ammonium (NH4 ug l-1) from 1988-1994. 


```{r}
#| warning: false
fig3
```



## References 

McDowell, William H., and USDA Forest Service. International Institute Of Tropical Forestry (IITF). 2024. “Chemistry of Stream Water from the Luquillo Mountains.” Environmental Data Initiative. [https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458](https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458).


Schaefer, Douglas. A., William H. McDowell, Fredrick N. Scatena, and Clyde E. Asbury. 2000. “Effects of Hurricane Disturbance on Stream Water Concentrations and Fluxes in Eight Tropical Forest Watersheds of the Luquillo Experimental Forest, Puerto Rico.” Journal of Tropical Ecology 16 (2): 189–207. [https://doi.org/10.1017/s0266467400001358.](https://doi.org/10.1017/s0266467400001358).


