--- 
title: "Time series of ion concentrations in Puerto Rico streams before and after Hurricane Hugo (1989)"
format: 
  html:
    theme: "pulse"
execute:
  message: false
  warning: false
---

![[Source](https://www.wbtw.com/news/grand-strand/remembering-hurricane-hugo-35-years-later/)](https://www.wbtw.com/wp-content/uploads/sites/22/2024/09/Screenshot-2024-09-21-080156.png?strip=1)

### Background 

Ion concentrations of potassium (K), nitrate (NO3-), magnesium (Mg), calcium (Ca), and ammonium (NH4+) in Puerto Rico Streams were collected from 1988-1984. After Hurricane Hugo hit Puerto Rico in Septermber of 1989, Shaefer et al. (2000) were interested in the hurricane's long-term impact on ion concentrations among four Puerto Rico streams in the Luquillo mountain region.  



### Data
Data is downloaded from [McDowell and International Institute Of Tropical Forestry (IITF) 2024](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064). 

### Methods
All .csv files are cleaned using 1_clean_data.R and moved into a output/output.csv. 1_clean_data.R filters individual .csv files for relevant dates (1988-1994) and selects only relevant columns (Sample_ID, Sample_Date, K, NO3-N, NH4-N, Mg, Ca). Then, moving_average() function, defined in R/moving-average.R, is applied to each individually subsetted data frame to calculate the 9-week moving averages. The resulting four dataframes are binded by rows and pivoted longer for faceting by ion. 

```{r}
#| output: false
# load package
library(tidyverse)
```

```{r}
# load cleaned data 
streams_pivot <- read_csv("../output/output.csv")
```

Use ggplot to plot data. 

Plot time on the x-axis, and mean ion concentration on the y-axis. Facet_wrap() by ion and stack them over each other. Add a vertical line on September 17, 1989 to represent date of Hurricane Hugo event. 


```{r}
# reorder the ions to match figure 3
streams_pivot$ion <- factor(
  streams_pivot$ion,
  levels = c("K", "NO3", "Mg", "Ca", "NH4")
)
```



```{r}
# plot
fig3 <- ggplot(
  streams_pivot,
  mapping = aes(
    x = window_start,
    y = mean_concentration,
    color = stream,
    linetype = stream
  )
  ) +
  geom_line() +
  facet_wrap(~ion, scales = "free", ncol = 1, strip.position = "left" ) +
  labs(
    x = "Date",
    y = "Ion concentration",
    caption = "Concentrations of ions before and after Hurricane Hugo in four Puerto Rico streams. "
    ) +
  geom_vline(xintercept = ymd("1989-09-17"), color = "darkgrey") +
  theme_bw()
```


# Results

Print the result.

```{r}
fig3
```