--- 
title: "Time series of ion concentrations in Puerto Rico streams before and after Hurricane Hugo (1989)"
format: html
---


# Background 

Ion concentrations of potassium (K), nitrate (NO3-), magnesium (Mg), calcium (Ca), and ammonium (NH4+) in Puerto Rico Streams were collected from 1988-1984. Hurricane Hugo hit Puerto Rico from Septermber 9-25, 1989, Shaefer et al. (2000) were interested in the hurricane's long-term impact on ion concentrations among four Puerto Rico streams in the Luquillo mountain region.  


# Data
Data is downloaded from [McDowell and International Institute Of Tropical Forestry (IITF) 2024](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064). 




# Methods

All .csv files are cleaned using 1_clean_data.R. This script filters for relevant dates (1988-1994) and selects only relevant columns (Sample_ID, Sample_Date, K, `NO3-N`, `NH4-N`, Mg, Ca). Then, moving_average(), defined in R/moving-average.R, is applied to each individually subsetted data to calculate the 9-week moving averages. The resulting four dataframes are binded by rows and pivoted longer for faceting by ion. 

Use ggplot to plot data. 

Plot time on the x-axis, and mean ion concentration on the y-axis. Facet_wrap() by ion and stack them over each other. Add a vertical line on September 17, 1989 to represent date of Hurricane Hugo event. 

```{r}

# load package
library(tidyverse)

# use cleaned data
streams_pivot <- read_csv("../output/output.csv")


# plot
figure_3 <- ggplot(
  streams_pivot,
  mapping = aes(
    x = window_start,
    y = mean_concentration,
    color = stream,
    linetype = stream
  )
) +
  geom_line() +
  facet_wrap(~ion, scales = "free", ncol = 1, ) +
  labs(
    x = "Date",
    y = "ion concentration",
    title = " Concentrations in PR Streams"
  ) +
  geom_vline(xintercept = ymd("1989-09-17"), color = "darkgrey") +
  theme_bw()
```


# Results

Plot the results.

```{r}
figure_3
```